---
documentclass: jss
author:
  - name: Alexander Rosenstock
    affiliation: |
      | Heinrich-Heine-Universität Düsseldorf \AND
      | ARAG SE
    address: |
      | Mathematisches Institut
      | Heinrich-Heine-Universität Düsseldorf
      | Universitätsstraße 1, 40225 Düsseldorf, Germany
    email: \email{alexander.rosenstock@hhu.de}
title:
  formatted: "Fitting Distributions and Neural Networks to Censored and Truncated Data: The \\proglang{R} Package \\pkg{reservr}"
  plain:     "Fitting Distributions and Neural Networks to Censored and Truncated Data: The R Package reservr"
  short:     "\\pkg{reservr}: Fitting Distributions and Neural Networks to Censored and Truncated Data"
abstract: >
  While the problem of estimating distribution parameters from truncated data with fixed truncation points has readily
  available implementations in \proglang{R}, this is not the case for randomly truncated data, i.e. observations where
  the point of truncation varies by observation.
  In practice, it is often of interest to also consider the influence of other parameters (predictors) on the
  distribution of an outcome variable.
  We present the features and implementation of the \proglang{R} package \pkg{reservr} for a large class of
  distributions.
  The package provides a flexible interface to specify distribution families, provides algorithms to perform parameter
  estimation for censored and randomly truncated data based on (conditional) maximum likelihood, and exposes an
  interface to the \proglang{R} package \pkg{tensorflow} for training flexible neural network models on censored and
  randomly truncated outcomes based on predictors.
  Additional utilities for application in a general insurance context, as well as the usual sampling, density,
  probability and quantile functions are provided.
keywords:
  formatted: [distribution fitting, truncation, random truncation, censoring, "\\proglang{R} packages"]
  plain:     [distribution fitting, truncation, random truncation, censoring, R packages]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
output: rticles::jss_article
bibliography: bibliography.bib
---

```{r, setup, include=FALSE}
library(reservr)
options(prompt = 'R> ', continue = '+ ')
```

# Introduction

Most statistical analyses are concerned with modelling and estimation of the distribution of some measured variable of interest, the outcome, possibly conditional on the value of one or several endogenous variables.
In the absence of endogenous variables, this process is usually called distribution fitting, and in the presence of endogenous variables it is called regression.
Classical regression, such as via generalized linear models, is usually concerned with the influence of endogenous variables on the mean of the outcome, i.e., $\mathbb{E}(Y|X) = f(X)$ for some random variable $Y$ and vector of associated endogenous variables $X$ paired with a homoskedascity assumption linking the other parameters of the outcome distribution to its mean.
Some models also allow specification of additional parameters of the distribution of the outcome variable, such as GAMLSS, which also allows for the scale parameter to depend on endogeneous variables.
More recently, deep distributional regression allows for flexible specification of individual outcome distribution parameters.
An example implementation of the latter can be found in \citet{deepregression}.

Usual statistical methods require complete data, that is full information on all observations of interest (X, Y).
In addition to this usual scenario we study two observation schemes that do not provide all observations.
Data with _interval censoring_ applied to the outcome $Y$ only consists of lower and upper bounds for $Y$ instead of the actual value.
Truncated data misses observations for which the outcome $Y$ falls out of a certain lower and upper truncation bound.
We consider the case of _random truncation_, where these truncation bounds are also random variables that vary for each observation.

[//]: # (TODO continue here)

Interval censored observations can be described as follows: let $(Y, M, V)$ denote a random vector, where $Y$ is the variable of interest that is supposed to have a density $f_\theta$ with respect to some dominating sigma-finite measuer $\mu$.
The pair $(M, V)$ shall satisfy $M \le Y \le V$, with $M$ possibly equal to $-\infty$ and $V$ possibly equal to $+\infty$.
Further, assume $Y$ to be conditionally independent of $(M, V)$ given $M \le Y \le V$ and $(M, V)$ to have density $f_{(M, V)}$ with respect to some dominating sigma-finite measure $\nu$.
A complete sample $(Y, M, V)$ is therefore drawn from the $(\mu \otimes \nu)$-density

\begin{align}
  f_{(Y, M, V) | M \le Y \le V}(y, m, v) & = f_{Y | M \le Y \le V}(y) f_{(M, V) | M \le Y \le V}(m, v) \nonumber\\
& = \frac{f_\theta(y)}{\mathrm{Pr}(M \le Y \le V)} f_{(M, V) | M \le Y \le V}(m, v) \mathbf{1}(m \le y \le v)
\end{align}


The random truncation problem can be formulated as follows: let $(Y, L, U)$ denote a random vector, where $Y$ is the variable of interest that is supposed to have a density $f_\theta$ with respect to some dominating sigma-finite measure $\mu$.
The pair $(L, U)$ is assumed to be independent of $Y$ and shall satisfy $L \le U$, with $L$ possibly equal to $-\infty$ and $U$ possibly equal to $+\infty$.
Further, $(L, U)$ shall have density $f_{(L, U)}$ with respect to some dominating sigma-finite measure $\nu$.
A sample of _randomly truncated observations_ from $(Y, L, U)$ consists of independent observations $(y_i, l_i, u_i)$ that we only happen to see if $l_i \le y_i \le u_i$.
As a consequence, any observed value can be regarded as being drawn from the $(\mu \otimes \nu)$-density

\begin{align}
  f_{(Y, L, U) | L \le Y \le U}(y, l, u) = \frac{f_{(L, U)}(l, u) f_\theta(y)}{\mathrm{Pr}(L \le Y \le U)} \mathbf{1}(l \le y \le u) \label{eq:trunc-dens}
\end{align}

Subsequently, we write $(Y^{(t)}, L^{(t)}, U^{(t)})$ for a random vector following the above density, i.e.,

\[
  f_{(Y^{(t)}, L^{(t)}, U^{(t)})}(y, l, u) = f_{(Y, L, U) | L \le Y \le U}(y, l, u)
\]

Estimating $\theta$ based on maximum likelihood requires specifying a distribution for $(L, U)$ (which can be regarded as a nuisance parameter) and calculating the denominator in \eqref{eq:trunc-dens}.
This (major) nuisance can be avoided by instead considering conditional maximum likelihood, which is known to produce consistent estimators as well.
In our case, we rely on considering the density of $Y^{(t)}$ conditional on the value of $(L^{(t)}, U^{(t)}) = (l, u)$, which is given by

\begin{align*}
  f_{Y^{(t)} | L^{(t)} = l, U^{(t)} = u}(y) & = \frac{f_{(Y^{(t)}, L^{(t)}, U^{(t)})}(y, l, u)}{f_{(L^{(t)}, U^{(t)})}(l, u)} \\
    & = \frac{f_{(Y, L, U) | L \le Y \le U}(y, l, u)}{\int_{[l, u]} f_{(Y, L, U) | L \le Y \le U}(z, l, u) \,\mathrm{d}z} = \frac{f_\theta(y)}{\int_{[l, u]} f_\theta(z) \,\mathrm{d}z}
\end{align*}

Combining interval censoring and random truncation yields the conditional likelihood

\begin{align}
  \ell(\theta; M^{(t)} = m, V^{(t)} = v, L^{(t)} = l, U^{(t)} = u) & = \log F_\theta([m, v]) - \log F_\theta([l, u])
\end{align}

A practical example of random truncation and interval censorship arises in the presence of inaccurate accident date specifications when reporting claims in general insurance.
The truncation bounds $(L, U)$ in this case will be equal to $(0, \tau - t_0)$ where $t_0$ is the actual accident time and $\tau$ is the calendar time.
Censoring bounds could, for example, be $(M, V) = (t_1 - \lceil t_0 \rceil, t_1 - \lfloor t_0 \rfloor)$ representing an inaccurate observation of $t_0$ at reporting time $t_1$.
Technically, $U$ is not (fully) observable in this situation, but can be approximated by $\tau - \lfloor t_0 \rfloor$.

For later purposes, it is helpful to attach a weight $w_i$ to each observation $(m_i, v_i, l_i, u_i)$ (one might think of $w_i = 1$ for the moment).
Denote the resulting sample by $\mathfrak{I} = \{(m_i, v_i, l_i, u_i, w_i)\}$ where the case $m_i = v_i$ denotes a non-censored observation and we have $l_i \le m_i \le v_i \le u_i$ for all observations.
We aim at maximizing

\begin{align}
  \ell(\theta | \mathfrak{I}) & = \sum_{(m, v, l, u) \in \mathfrak{I}} w \cdot \begin{cases}
    \log f_\theta(m) - \log F_\theta ([l, u]) & m = v \\
    \log F_\theta([m, v]) - \log F_\theta([l, u]) & m < v
  \end{cases} \label{eq:cml-likelihood}
\end{align}

## Related packages

For the less general cases of censoring without truncation and truncation without censoring, as well as for estimation of distribution parameters in the absence of any censoring or truncation mechanism, there are a number of \proglang{R} packages that can fit distributions, some of them also supporting weights.
Among these are \pkg{MASS} \citep{MASS}, \pkg{fitdistrplus} \citep{fitdistrplus}, \pkg{evmix} \citep{evmix}, \pkg{ExtDist} \citep{ExtDist}, \pkg{flexsurv} \citep{flexsurv}.
Another R6-based interface is provided by \pkg{ROOPSD} \citep{ROOPSD}.

The remaining parts of this paper are structured as follows: Section&nbsp;\ref{pkg-overview} details the core functionality of the corresponding \proglang{R} package \pkg{reservr}.
It is split into definition of samples $\mathfrak{I}$ (Section&nbsp;\ref{trunc-obs}), definition of distribution families (Section&nbsp;\ref{distributions}), mathematical definitions of some available distribution families (Section&nbsp;\ref{dist-definitions}), estimation of distribution parameters (Section&nbsp;\ref{fit-dist}) and distributional regression using \pkg{tensorflow} (Section&nbsp;\ref{tensorflow}).
A conclusion is given in Section&nbsp;\ref{conclusion}.

# Usage of \pkg{reservr} {short-title="Usage of reservr" #pkg-overview}

The package serves two main goals: fitting distributions to randomly truncated interval censored data and performing (deep) distributional regression with randomly truncated interval censored data.
Four main components are integrated with each other to facilitate the analysis goals

1. Methods for defining a randomly truncated interval censored sample $\mathfrak{I}$.
2. Methods for specifying distribution families $\mathcal{F}$ to be fitted.
3. Methods for estimating distribution parameters given a sample $\mathfrak{I}$ and a family $\mathcal{F}_\theta$.
4. Methods for regression of distribution parameters given a sample $\mathfrak{I}$ with associated covariates $X$, a family $\mathcal{F}_\theta$ and a general \pkg{tensorflow} network $g$ that processes $X$ to estimate $Y | X = F_{g(x)}$.

Each of these components is described one by one in the following sections.

## Working with samples {#trunc-obs}

A sample $\mathfrak{I} = \{(m, v, l, u, w)_i\}$ is represented as a tibble (from package \pkg{tibble}).
The core function to create this tibble is `trunc_obs()`.
A tibble created by `trunc_obs()` consists of five columns:

 * `x`: If observed, the exact value of the random variable. Otherwise `NA`.
 * `xmin`: Lower interval censoring bound for the observation. If the observation is not censored, `xmin` is equal to `x`.
 * `xmax`: Upper interval censoring bound for the observation. If the observation is not censored, `xmax` is equal to `x`.
 * `tmin`: Lower truncation bound. Only observations with $x \ge t_{\text{min}}$ are observed. Can be $-\infty$ to indicate no lower truncation.
 * `tmax`: Upper truncation bound. Only observations with $x \le t_{\text{max}}$ are obserbed. Can be $\infty$ to indicate no upper truncation.
 * `w`: The weight associated with the observation. Defaults to $1$.

The following code defines a sample of size 1 without truncation and censoring, with the realized value of $1.3$.

```{r}
trunc_obs(1.3)
```

Simulating randomly truncated and interval censored data from a standard normal distribution with $80\%$ of the observations randomly interval censored and random uniform truncation $L \sim U[-2, 0]$ and $U \sim U[0, 2]$ can be simulated as follows

```{r}
set.seed(123)
N <- 1000L
x <- rnorm(N)
is_censored <- rbinom(N, size = 1L, prob = 0.8) == 1L
x_lower <- x
x_lower[is_censored] <- x[is_censored] -
  runif(sum(is_censored), min = 0, max = 0.5)
x_upper <- x
x_upper[is_censored] <- x[is_censored] +
  runif(sum(is_censored), min = 0, max = 0.5)

t_lower <- runif(N, min = -2, max = 0)
t_upper <- runif(N, min = 0, max = 2)

is_observed <- t_lower <= x & x <= t_upper

obs <- trunc_obs(
  xmin = pmax(x_lower, t_lower)[is_observed],
  xmax = pmin(x_upper, t_upper)[is_observed],
  tmin = t_lower[is_observed],
  tmax = t_upper[is_observed]
)
```

Observations look like:

```{r}
obs[8L:12L, ]
```

The total number of observations is smaller than the base population of $1000$ due to truncation:
```{r}
nrow(obs)
```

The total number of censored observations is roughly $0.8 \cdot \mathtt{nrow(obs)}$.
```{r}
sum(is.na(obs$x))
```

In addition to the `trunc_obs()` constructor function, there are `as_trunc_obs()` for coercion, `truncate_obs()` for artifically changing truncation bounds, and `repdel_obs()` for computing randomly truncated reporting delay observations from general insurance claims data containing accident date, reporting delay and evaluation date information.
The latter takes inputs of the form $(T_\text{acc}, D, \tau)$ and returns the sample $(\mathtt{xmin} = \mathtt{xmax} = D, \mathtt{tmin} = 0, \mathtt{tmax} = \tau - T_{\text{acc}}, \mathtt{w} = 1)$ suitable for estimating the reporting delay distribution where a claim is only observed if it has been reported by the evaluation date, i.e., $T_{\text{acc}} + D \le \tau$.

## Definition of distribution families {#distributions}

Distribution families are implemented using the \pkg{R6} class system \citep{R6}.
They inherit from `Distribution` and feature a common interface to

* Manage fixed and free parameters of the underlying familiy
* Basic distribution functions for sampling, computation of density, cumulative density, hazard and quantile
* Additional functions supporting parameter estimation procedures such as computing support and discreteness
* Performance enhancements to speed up basic functions for repeated evaluation
* \pkg{tensorflow}-specific implementations to support (deep) distributional regression

A `Distribution` object represents a family $\mathcal{F}$ parameterized by a fixed-dimensional parameter space $\Theta$.
This implies that, e.g., for mixture distributions, the total number of component distributions and their families needs to be fixed.

\pkg{reservr} provides a set of basic distributions with fixed parameters as well as some transformations of distributions that take one or more underlying distributions.
At the time of writing, these are:

| Generator function      &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | Description                                                                                             |
|--------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------|
| `dist_bdegp(n, m, u, epsilon)`                                           | A Blended Dirac Erlang Generalized Pareto distribution, see Section&nbsp;\ref{dist-bdegp}               |
| `dist_beta(shape1, shape2, ncp)`                                         | A (non-central) Beta distribution                                                                       |
| `dist_binomial(size, prob)`                                              | A Binomial distribution                                                                                 |
| `dist_blended(dists, probs, breaks, bandwidths)`                         | A Blended mixture distribution, see Section&nbsp;\ref{dist-blended}                                     |
| `dist_dirac(point)`                                                      | A Dirac distribution with full mass at `point`                                                          |
| `dist_discrete(size, probs)`                                             | A discrete distribution with support $\{1, \ldots, \text{size}\}$ and $P(X = k) = \text{probs}_k$       |
| `dist_empirical(sample, positive, bw)`                                   | An empirical distribution derived from a sample, with density estimated using a kernel density estimate |
| `dist_erlangmix(shapes, scale, probs)`                                   | An Erlang mixture distribution, see Section&nbsp;\ref{dist-erlangmix}                                   |
| `dist_exponential(rate)`                                                 | An Exponential distribution                                                                             |
| `dist_gamma(shape, rate)`                                                | A Gamma distribution                                                                                    |
| `dist_genpareto(u, sigmau, xi)`                                          | A Generalized Pareto Distribution                                                                       |
| `dist_genpareto1(u, sigmau, xi)`                                         | A Generalized Pareto Distribution with the tail index $\xi$ constrained to $(0, 1)$                     |
| `dist_lognormal(meanlog, sdlog)`                                         | A Log-Normaldistribution                                                                                |
| `dist_mixture(dists, probs)`                                             | A general Mixture distribution, see Section&nbsp;\ref{dist-mixture}                                     |
| `dist_negbinomial(size, mu)`                                             | A negative Binomial distribution                                                                        |
| `dist_normal(mean, sd)`                                                  | A Normal distribution                                                                                   |
| `dist_pareto(shape, scale)`                                              | A Pareto Type I distribution, i.e., a Generalized Pareto distribution with `u` = $0$                    |
| `dist_poisson(lambda)`                                                   | A Poisson distribution                                                                                  |
| `dist_translate(dist, offset, multiplier)`                               | An affine transformation $Y = a \cdot X + b$ of another distribution                                    |
| `dist_trunc(dist, min, max)`                                             | A truncated distribution, $Y = X \mid \text{min} \le X \le \text{max}$                                  |
| `dist_uniform(min, max)`                                                 | A uniform distribution                                                                                  |
| `dist_weibull(shape, scale)`                                             | A Weibull distribution                                                                                  |

### Parameters

Parameters of distribution families can either be fixed to a set value, or free.
Free parameters (_placeholders_) are those that should be estimated from data whereas fixed parameters are held fixed.
Most distribution methods have an argument `with_params` to provide values for the free parameters and need fully specified parameters to work.
For example, generating samples from a distribution is only possible if it is fully parameterized using fixed parameters and the `with_params` argument of `Distribution$sample()`.

Distributions have a set of fields and methods related to managing parameters

* `default_params`, `get_params()` gets or sets the list of all parameters and their fixed values, `NULL` represents a free parameter. `get_params()` traverses component distribution parameters while `default_params` contains the `Distribution` objects for the respective parameters.
* `get_placeholders()` gets the list of free parameters with `NULL` as values.
* `param_bounds`, `get_param_bounds()` gets or sets the domain of each parameter as an `Interval` object. Setting a bound via the `param_bounds` active binding allows restricting the natural parameter space of a distribution. `get_param_bounds()` only returns the bounds of free parameters while `param_bounds` includes all parameters.
* `get_param_constraint()` returns `NULL` or a function that evaluates non-box constraints on the parameter set. The function must return a list with elements `constraints` and `jacobian` or just a vector of constraint values (that need to be equal to $0$ for valid parameters), optionally together with the jacobian of the constraint function. Used in `nloptr::slsqp(heq=)` for estimation. An example is that mixture distributions require the `probs` parameters to sum to $1$ in addition to the box constraint that each parameter is in $[0, 1]$.
* `get_components()` returns a list of component distributions for transformations or mixtures. The list is empty for basic families.

Here is an example for a normal family with fixed standard deviation $\sigma = 1$ and a mixture distribution with two components, one of which is specified as a normal distribution:

```{r}
dist <- dist_normal(sd = 1.0)
mix <- dist_mixture(dists = list(dist_normal(), NULL))

dist$default_params
mix$default_params
str(dist$get_placeholders())
str(mix$get_placeholders())
str(dist$param_bounds)
str(mix$param_bounds)
str(dist$get_param_bounds())
str(mix$get_param_bounds())
str(dist$get_param_constraints())
str(mix$get_param_constraints())
dist$get_components()
mix$get_components()
```

### Basic distribution functions

The basic distribution functions (density, probability, hazard, quantile, sample) are provided by each distribution with signatures.
In general, the argument `with_params` can be used to both specify missing parameters (placeholders) and to override fixed distribution parameters.
If the provided parameters are vectors of length greater than 1, they must conform to the input dimension (e.g. `x` for `density`).
In this case, the parameters are "vectorized" in the sense that the $i$th output element will be computed using the $i$th entry from the parameter list.

 * `density(x, log = FALSE, with_params = list())` compute (log-)density.
 * `probability(q, lower.tail = TRUE, log.p = FALSE, with_params = list()` compute (log-)cumulative density or (log-)survival.
 * `hazard(x, log = FALSE. with_params = list())` compute (log-)hazard.
 * `quantile(p, lower.tail = TRUE, log.p = FALSE, with_params = list())` compute upper or lower quantiles.
 * `sample(n, with_params = list())` generate a random sample (`with_params` can contain length `n` vectors in this case).

### Additional functions

In addition to the basic functions, there are several supporting functions useful for faster or more accurate estimation of parameters.
These functions are not necessarily implemented for all distribution classes, but will be automatically used by, e.g., `fit_dist()` if useful.

 * `export_functions(name, with_params = list())` exports `{d,p,q,r}<name>` functions adhering to the common R convention for distribution functions.
 * `get_type()` returns one of `"continuous"`, `"discrete"`, or `"mixed"` depending on whether the distribution has a density with respect to the Lebesgue measure, the counting measure, or their sum.
 * `is_continuous()` and `is_discrete()` testing for the particular type.
 * `has_capability(caps)` gives information on whether a specific implementation provides some or all of the features described.
   Possible capabilities are `"sample"`, `"density"`, `"probability"`, `"quantile"`, `"diff_density"`, `"diff_probability"`, `"tf_logdensity"`, `"tf_logprobability"`.
 * `require_capability(caps)` errors if the specified capabilities are not implemented for the distribution at hand.
 * `is_discrete_at(x, with_params = list())` returns a logical vector indicating whether the distribution has a point mass at `x`.
 * `is_in_support(x, with_params = list())` returns a logical vector indicating whether the distribution has any mass at `x`.

### Performance enhancements

When working with larger data or many calls to distribution functions, such as when performing a fit, it can be beneficial to just-in-time compile specialized functions that avoid overhead for dealing with the generic structure of distributions and their parametrization.
Distributions offer a set of "compiler" functions that return simplified, faster, versions of the basic distribution functions, or that avoid having to numerically compute gradients where possible.

 * `compile_density()` compiles a fast function with signature `(x, param_matrix, log = FALSE)` that will compute the density with fixed parameters hard-coded and taking the free parameters as a matrix with defined layout instead of a nested list.
 * `compile_probability()` compiles a fast replacement for `probability` with signature `(q, param_matrix, lower.tail = TRUE, log.p = FALSE)`.
 * `compile_probability_interval()` compiles a fast function with signature `(qmin, qmax, param_matrix, log.p = FALSE)` computing $P(X \in [\mathtt{qmin}, \mathtt{qmax}])$ or its logarithm efficiently. This expression is necessary for computing truncation probabilities.
 * `compile_sample()` compiles a fast replacement for `sample` with signature `(n, param_matrix)`.
 * `diff_density(x, log = FALSE, with_params = list())` computes the (log-)gradients of the density function with respect to free distribution parameters, useful for maximum likelihood optimization.
 * `diff_probability(q, lower.tail = TRUE, log.p = FALSE, with_params = list())` computes the (log-)gradients of the cumulative density function with respect to free distribution parameters

### \pkg{tensorflow} interface {short-title="tensorflow interface"}

Use of distributions from within \pkg{tensorflow} networks requires specialized implementations using the \pkg{tensorflow} APIs instead of regular \proglang{R} functions.
These are tailored to the needs of maximizing (conditional) likelihoods of weighted, censored and randomly truncated data.

 * `tf_compile_params(input, name_prefix = "")` creates \pkg{keras} layers that take an `input` layer and transform it into a valid parametrization of the distribution.
 * `tf_is_discrete_at()` returns a \pkg{tensorflow}-ready version of `is_discrete_at()`.
 * `tf_logdensity()` returns a \pkg{tensorflow}-ready version of  `compile_density()` with implied `log = TRUE`.
 * `tf_logprobability()` returns a \pkg{tensorflow}-ready version pf `compile_probability_interval()` with implied `log.p = TRUE`.
 * `tf_make_constants()` creates a list of constant tensors for all fixed distribution parameters.

## Special families {#dist-definitions}

Some of the distribution families available in \pkg{reservr} have tailored algorithms for estimation, or are not commonly known.
This section contains mathematical definitions of those function families.

### Mixture distribution families {#dist-mixture}

A mixture distribution family is defined by a fixed number $k$ of component families via the set of distributions

\begin{align*}
  \mathop{\mathrm{Mixture}}(\mathcal{F}_1, \ldots, \mathcal{F}_k) & := \{ F \mid F((-\infty, x]) = \sum_{i = 1}^k p_i F_i((-\infty, x]), F_i \in \mathcal{F}_i, p_i \in [0, 1], \sum_{i = 1}^k p_i = 1 \}
\end{align*}

### Erlang mixture distribution families {#dist-erlangmix}

An Erlang mixture distribution family is defined by its number of components $k$ as a mixture of Erlang distributions (Gamma distributions with integer shape parameter) with common scale parameter.
If $\Gamma_{\alpha, \theta}$ denotes a Gamma distribution with shape $\alpha$ and scale $\theta$, the erlang mixture family with $k$ components can be defined as follows:

\begin{align*}
  \mathop{\mathrm{ErlangMixture}}(k) := \Bigl\{ F \mid & F((-\infty, x]) = \sum_{i = 1}^k p_i \Gamma_{\alpha_i, \theta}((-\infty, x]), \\
  & \alpha_i \in \mathbb{N}, \theta \in (0, \infty), p_i \in [0, 1], \sum_{i = 1}^k p_i = 1 \Bigr\}.
\end{align*}

Note that for $k \to \infty$, Erlang mixtures are dense in the space of distributions on $(0, \infty)$, making them a useful modeling choice for general positive continuous distributions. However, the tail index of all Erlang mixture distributions is always zero due to the exponential decay of Gamma densities.

### Blended distribution families {#dist-blended}

A Blended distribution is defined in \citet{Rosenstock2022} as follows: Given two underlying distributions $P, Q$ on $\mathbb{R}$ with cdfs $F(\cdot)=P((-\infty, \cdot])$ and $G(\cdot)=Q((-\infty, \cdot])$, respectively, and parameters $\kappa \in \mathbb{R}, \varepsilon \in (0, \infty), p \in [0, 1]^2, p_1 + p_2 = 1$ such that $F(\kappa) > 0$ and $G(\kappa) < 1$, we define the \textit{Blended Distribution} $B = \mathop{\mathrm{Blended}}(P, Q; p, \kappa, \varepsilon)$ of $P$ and $Q$ with blending interval $[\kappa - \varepsilon, \kappa + \varepsilon]$ and mixture probabilities $p$ via its cdf
$F_B$:
\begin{align*}
    p_{\kappa, \varepsilon}(x) &=
    \begin{cases}
        x & , x \in (-\infty, \kappa-\varepsilon],\\
        \tfrac12 (x + \kappa - \varepsilon) +
        \tfrac\varepsilon\pi \cos\Big( \frac{\pi (x - \kappa)}{2 \varepsilon} \Big) &, x \in (\kappa-\varepsilon , \kappa+\varepsilon], \\
        \kappa &, x  \in (\kappa +\varepsilon, \infty),
    \end{cases} \\
    \nonumber
    q_{\kappa, \varepsilon}(x) & =
    \begin{cases}
        \kappa & , x \in (-\infty, \kappa-\varepsilon],\\
        \tfrac12 (x + \kappa + \varepsilon) -
        \tfrac\varepsilon\pi \cos\Big( \frac{\pi (x - \kappa)}{2 \varepsilon} \Big) &, x \in (\kappa-\varepsilon , \kappa+\varepsilon], \\
        x &, x  \in (\kappa +\varepsilon, \infty),
    \end{cases} \\
    F_B(x) & = p_1 \frac{F(p_{\kappa, \varepsilon}(x))}{F(\kappa)} + p_2 \frac{G(q_{\kappa, \varepsilon}(x)) - G(\kappa)}{1 - G(\kappa)}.
\end{align*}

This leads to the definition of a blended distribution family:

Given two families $\mathcal{F}, \mathcal{G}$ of distributions on $\mathbb{R}$, and parameters $\kappa \in \mathbb{R}, \varepsilon \in (0, \infty)$, we define the \textit{Blended Distribution family} as the family of Distributions
\begin{align*}
    \mathop{\mathrm{Blended}}(\mathcal{F}, \mathcal{G}; \kappa, \varepsilon) & := \{ \mathop{\mathrm{Blended}}(P, Q ; p, \kappa, \varepsilon) \mid P \in \mathcal{F}, Q \in \mathcal{G}, p \in [0, 1]^2, \lVert p\rVert_1 = 1 \}.
\end{align*}

Blended distribution families can be generalized to a number of components $k$ by letting $\kappa$ and $\varepsilon$ become vectors of dimension $k - 1$ such that $\kappa_i + \varepsilon_i \le \kappa_{i + 1} - \varepsilon_{i + 1}$ for $i = 1, \ldots, k - 2$.
Compared to piecewise distribution families obtained by mixture of truncated distribution families with supports $(-\infty, \kappa]$ and $[\kappa, \infty)$ such as those commonly used for extreme value modelling, blended distribution families exhibit a continuous density within the blending region $(\kappa - \varepsilon, \kappa + \varepsilon)$.

\pkg{reservr} provides an implementation via `dist_blended()`.

### The Blended Dirac Erlang Generalized Pareto distribution family {#dist-bdegp}

Using the construction of a Blended distribution family, we can define the Blended Dirac Erlang Generalized Pareto (BDEGP) family as follows:

Given parameters $n \in \mathbb{N}, m \in \mathbb{N}, \kappa \in \mathbb{R}$ and $\varepsilon \in (0, \infty)$, we define the \textit{Blended Dirac Erlang Generalized Pareto family} as the family of distributions

\begin{align*}
  \mathop{\mathrm{BDEGP}}(n, m, \kappa, \varepsilon) := & \mathop{\mathrm{Mixture}}( \\
    & \qquad \{\delta_0\}, \{\delta_1\}, \ldots, \{\delta_{n-1}\}, \\
    & \qquad \mathop{\mathrm{Blended}}( \\
      & \qquad\qquad \mathop{\mathrm{ErlangMixture}}(m), \\
      & \qquad\qquad \{ \mathop{\mathrm{GPD}}(\kappa, \sigma, \xi) \mid \sigma \in (0, \infty), \xi \in [0, 1]) \}; \\
      & \qquad\qquad \kappa, \varepsilon \\
    & \qquad) \\
  &)
\end{align*}

Where $\delta_k$ is the dirac distribution with $P(X = k) = 1$ and $\mathrm{GPD}$ is the generalized Pareto distribution. Note the constraint on the tail index $\xi \in [0, 1]$.

This distribution family has three features making it useful in modelling very general heavy-tailed distributions on $(0, \infty)$:

1. A maximally flexible lower tail
2. A flexible family of distributions for its body
3. A flexible tail index due to the generalized Pareto component

## Methods of estimating distribution parameters {#fit-dist}

This section describes the functions for the problem of estimating a parameter $\theta \in \Theta$ given a sample $\mathfrak{I}$ and a parameterized family $\mathcal{F} = \{F_\theta \mid \theta \in \Theta\}$.
Sometimes, the conditional likelihood in \eqref{eq:cml-likelihood} can be directly maximized, yielding an estimate for $\theta$.
This is the default behavior in \pkg{reservr} if no specialized estimation routine for the provided family $\mathcal{F}_\theta$ is found.
Depending on whether there are box constraints, nonlinear constraints or no constraints on the parameter space $\Theta$, different implementations of nonlinear optimization algorithms from \pkg{nloptr} \citep{nloptr,nloptr-tnewton,nloptr-lbfgs,nloptr-slsqp} are employed.

In addition to the naive direct optimization approach, some families lend themselves to specialized estimation algorithms which usually show faster convergence due to making use of special structures in the parameter space $\Theta$.

Fitting distributions to truncated observations is handled using the generic `fit()` method.
It delegates to `fit_dist()`, which is also generic with signature:

* `dist`: The distribution family to be fit
* `obs`: The `trunc_obs` object, or a vector of observed values
* `start`: Starting parameters, as a list compatible with `dist$get_placeholders()`.

At the time of writing there are specialized algorithms for three types of families:

1. Blended distributions \citep[Algorithm 2]{Rosenstock2022}
2. Erlang mixture distributions (GEM-CMM algorithm from \citet{Gui2018}, with a fixed number of components $M$)
3. Generalized pareto distributions with free lower bound `u`
4. Mixture distributions \citep[Algorithm 1]{Rosenstock2022}
5. Translated distributions with fixed `offset` and `multiplier` (transform the sample to the space of the component distribution family before fitting)
6. Uniform distributions (match free parameters to the range of the sample directly)

If not present, the `start` parameter is obtained via the `fit_dist_start()` generic.
This generic implements a family specific method of generating valid starting values for all placeholder parameters.
A notable implementation is `fit_dist_start.ErlangMixtureDistribution()` for Erlang mixture distributions.
If the shape parameters are free, there are different initialization strategies that can be chosen using additional arguments to `fit_dist_start()`:

* `init = "shapes"` paired with `shapes = c(...)` manually specifies starting shape parameters $\alpha$
* `init = "fan"` paired with `spread = d` uses $\alpha = (1, 1 + d, \ldots, 1 + (k - 1) \cdot d)$ with a default of $d = 1$ resulting in $\alpha = (1, \ldots, k)$
* `init = "kmeans"` uses 1-dimensional K-means based clustering of the sample observations such that each cluster corresponds to a unique shape
* `init = "cmm"` uses the centralized method of moments procedure described in \citet{Gui2018}

Re-using `dist` $= \mathcal{N}_{\sigma = 1}$ from above and the generated sample `obs`, we can fit the free parameter `mean`:

```{r}
the_fit <- fit(dist, obs)
str(the_fit)
```

Using the function `plot_distributions()` we can also assess the quality of the fit.

```{r}
plot_distributions(
  true = dist,
  fitted = dist,
  empirical = dist_empirical(0.5 * (obs$xmin + obs$xmax)),
  .x = seq(-5, 5, length.out = 201),
  plots = "density",
  with_params = list(
    true = list(mean = 0.0, sd = 1.0),
    fitted = the_fit$params
  )
)
```

We follow with an example of fitting an $\mathrm{ErlangMixture}(3)$ distribution family using various initialization strategies.
Note that both, `"kmeans"` and `"cmm"` use the random number generator for internal K-means clustering.
This necessitates setting a constant seed before running `fit_dist_start()` and `fit()` to ensure the chosen starting parameters are the same for both calls.

```{r}
dist <- dist_erlangmix(list(NULL, NULL, NULL))
params <- list(
  shapes = list(1L, 4L, 12L),
  scale = 2.0,
  probs = list(0.5, 0.3, 0.2)
)

set.seed(1234)
x <- dist$sample(100L, with_params = params)

set.seed(32)
init_true <- fit_dist_start(dist, x, init = "shapes",
                              shapes = as.numeric(params$shapes))
init_fan <- fit_dist_start(dist, x, init = "fan", spread = 3L)
init_kmeans <- fit_dist_start(dist, x, init = "kmeans")
init_cmm <- fit_dist_start(dist, x, init = "cmm")
flatten_params(init_true)
flatten_params(init_fan)
flatten_params(init_kmeans)
flatten_params(init_cmm)

set.seed(32)
str(fit(dist, x, init = "shapes", shapes = as.numeric(params$shapes)))
fit(dist, x, init = "fan", spread = 3L)$logLik
fit(dist, x, init = "kmeans")$logLik
fit(dist, x, init = "cmm")$logLik
```

It should be noted that the different initialization methods had a considerable impact on the outcome due to the discrete nature of Erlang mixture distributions and thus the combinatorial difficulty of picking optimal shapes $\alpha$.
The `fit()` result for Erlang mixture distributions contains an element named `"params_hist"`.
This can be populated by passing `trace = TRUE` to `fit()` and will record parameters after all ECME steps.
The element `"iter"` contains the number of full ECME-Iterations that were performed.

## Distributional regression using \pkg{tensorflow} integration {short-title="distributional regression using tensorflow integration" #tensorflow}

This section describes the functions for the problem of regressing a function $g : \mathfrak{X} \to \Theta$ given a sample $\mathfrak{I}$ with associated covariates $X \in \mathfrak{X}$, a parameterized family $\mathcal{F} = \{F_\theta \mid \theta \in \Theta\}$ and a family $\mathcal{G}$ of functions from $\mathfrak{X}$ to $\Theta$.
Distributional regression can be formulated as

\begin{align}\label{eq:cml-regression-likelihood}
  \ell(g | \mathfrak{I}, X) & := \sum_{(m, v, l, u) \in \mathfrak{I}, x \in X} w \cdot \begin{cases}
    \log f_{g(x)}(m) - \log F_{g(x)} ([l, u]) & m = v \\
    \log F_{g(x)}([m, v]) - \log F_{g(x)}([l, u]) & m < v
  \end{cases}, \text{and} \\
  \hat g & = \mathop{\mathrm{arg\,max}}_{g \in \mathcal{G}} \ell(g | \mathfrak{I}, X). \nonumber
\end{align}

Compared with \eqref{eq:cml-likelihood}, the global parameter $\theta$ is replaced by the regression function $g$ evaluated at the associated covariates.

The maximization problem is delegated to \pkg{tensorflow}, which supplies ample stochastic optimization algorithms.
Functions in \pkg{reservr} are necessary to create a suitable output layer for \pkg{tensorflow} that maps onto $\Theta$ and to provide an implementation of the (negative) log-likelihood \eqref{eq:cml-regression-likelihood}.
These two tasks are combined in `tf_compile_model()`.
The function returns an object of class `reservr_keras_model`, which can be used for the estimation procedure.

Given input layers `inputs` and an intermediate output layer `intermediate_output` as well as a family of distributions `dist`, the function

* Compiles the loss for `dist` defined by \eqref{eq:cml-regression-likelihood} as $l(g) = -\tfrac{1}{\#(\mathfrak{I})} \ell(g|\mathfrak{I}, X)$, optionally disabling `censoring` or `truncation` for efficiency.
* Creates a list of final output layers mapping `intermediate_output` onto the parameter space $\Theta$ of `dist` using `Distribution$tf_compile_params()`.
* Runs `keras::compile()` on the underlying `keras.engine.training.Model`.

The following example defines a linear model with homoskedasticity assumption and fits it using $100$ iterations of the Adam optimization algorithm.
First, we simulate data from the model $Y | X = \mathcal{N}(\mu 2 \cdot X, \sigma = 1)$.

```{r}
set.seed(1431L)
tensorflow::set_random_seed(1432L)

dataset <- tibble::tibble(
  x = runif(100, min = 10, max = 20),
  y = 2 * x + rnorm(100)
)
```

Next, we specify the distribution family $\mathcal{F} = \mathcal{N}_{\sigma = 1}$, incorporating the homoskedasticity assumption.

```{r}
dist <- dist_normal(sd = 1.0)
```

Using \pkg{keras}, we define an empty neural network, just taking $X$ as an input and performing no transformation.

```{r}
nnet_input <- keras::layer_input(shape = 1L, name = "x_input")
nnet_output <- nnet_input
```

Then, `tf_compile_model()` adapts the input layer to the free parameter space $\Theta = \mathbb{R}$.
This introduces two parameters to the function family $\mathcal{G}$ and implies the functional relationship $\mu = \theta_1 \cdot x + \theta_0$.
Since our sample is fully observed, we disable censoring and truncation, leading to the simplified loss

\begin{align*}
  l(g) = -\tfrac{1}{100} \sum_{x, y} \log f_{g(x)}(y),
\end{align*}

where $f_\mu(y)$ is the density of $\mathcal{N}_{\mu=\mu, \sigma=1}$ evaluated at $y$.

```{r}
nnet <- tf_compile_model(
  inputs = list(nnet_input),
  intermediate_output = nnet_output,
  dist = dist,
  optimizer = keras::optimizer_adam(learning_rate = 0.1),
  censoring = FALSE,
  truncation = FALSE
)
nnet$dist
nnet$model
```

The fit can now be performed, modifying the parameters (weights) of `nnet` in-place.
Note that the argument `y` of fit accepts a `trunc_obs` object.
In our example, the vector `y` is silently converted to an untruncated, uncensored `trunc_obs` object.
`fit()` returns the `keras_training_history` of the underlying call to `fit()` on the `keras.engine.training.Model`.

```{r}
nnet_fit <- fit(
  nnet,
  x = dataset$x,
  y = dataset$y,
  epochs = 100L,
  batch_size = 100L,
  shuffle = FALSE,
  verbose = FALSE
)
plot(nnet_fit)
```

The `predict()` method of `reservr_keras_model` takes input tensors and returns the predicted distribution parameters as a list compatible with `dist$get_placeholders()`.
We can thus extract the only parameter `mean` and compare it to an OLS fit for the same dataset:

```{r}
pred_params <- predict(nnet, data = list(keras::k_constant(dataset$x)))

lm_fit <- lm(y ~ x, data = dataset)

dataset$y_pred <- pred_params$mean
dataset$y_lm <- predict(lm_fit, newdata = dataset, type = "response")

library(ggplot2)
ggplot(dataset, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(y = y_pred), color = "blue") +
  geom_line(aes(y = y_lm), linetype = 2L, color = "green")
```

Since a `reservr_keras_model` includes the underlying `keras.engine.training.Model`, its parameters can also be extracted and compared to the OLS coefficients

```{r}
coef_nnet <- rev(as.numeric(nnet$model$get_weights()))
coef_lm <- unname(coef(lm_fit))

str(coef_nnet)
str(coef_lm)
```

# Conclusion {#conclusion}

TODO conclude